One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2021-09-02 paper.tex
edits up to 3.2

Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and an assumed detection model to estimate the detectability of animals as a function of distance from transect \citep{buckland_advanced_2004, buckland_distance_2015}.  Classic distance sampling approaches use a hybrid of design- and model-based inference to estimate population size where the probability of detection is modelled and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.

More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow non-randomized survey designs to be used.  These methods allow density to be associated with spatially-indexed covariates and animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.  This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.

Model-based distance sampling has most commonly been implemented in a two-stage modelling framework, see the \texttt{dsm} package \citep{miller_spatial_2013}, for example.  In the first stage, detectability is estimated.  In the second stage, detectability estimates from the first stage are used as an offset in a generalized additive model framework.  Detections within sampling units are binned into counts and an appropriate response distribution is chosen.  Due to the often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.  Negative-binomial and Tweedie distributions are common choices.  This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review.

A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the second-stage spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_reliable_2018-1, williams_chilean_2011}. Instead, \cite{bravington_reliable_2018-1} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  These approaches tend to be Bayesian and use data augmentation to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference \citep{royle_hierarchical_2008, schmidt_using_2012}.  \citet{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.

The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use the approximate Bayesian method of integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} applied to line transect distance sampling data.  INLA is a popular choice for fitting models with spatially-structured random effects in large or complex spatial domains due to the implementation of a computationally efficient approach based on the construction of a sparse prior precision matrix for the random effect parameters.  This sparse precision matrix specifies a GMRF representation of a Gaussian random field defined as the stationary solutions to an SPDE \citep{lindgren_explicit_2011} (see Section \ref{sec-gmrf} for more details on this approach).  The finite element methods involved use a flexible triangulation of the spatial domain and have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, a useful feature for applications with complex boundaries which appear in wildlife survey data such as boundaries formed by coastlines, rivers and other natural features.

\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of the full, but imperfectly observed, point pattern.  Key to their approach was formulating the detection model as the solution to an SPDE.  Choosing a B-spline basis and applying finite element methods results in a sparse prior precision matrix for the detection model coefficients which can then be incorporated into INLA using the same machinery as already exists for latent Gaussian precision matrices.  A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing - usually a key feature of detection functions.  The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.

Here we present a related approach to the one in \citet{yuan_point_2017} in that our models are fitted using INLA.  However, we avoid these issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal, as a user-written R function.  This approach of choosing a parametric family of detection functions will be more familiar to users of classical distance sampling methods than the SPDE specification.  However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \text{R-INLA} \citep{rue_approximate_2009} package implementation of INLA.  We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.

We apply these methods to point transect \akepa data which differ from line transect methods to account for the fact that the area surveyed increases with increasing distance from the transect.  Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.

>>>>

Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_advanced_2004, buckland_distance_2015}.  Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.  The probability of detection is modelled using a parametric equation and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.  Under this approach, animal density is uniform within each strata of the survey.

More recently, interest has focused on fully model-based approaches that model a spatially varying animal density and allow for the use of non-random survey designs.  These methods allow animal density to be associated with spatially-indexed covariates and non-parametric random effects.  Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.  As such, varying density estimates are not restricted by the stratification design required for Horvitz-Thompson-like estimators.  This additional flexibility to make predictions for smaller, user-specified geographic units, and extrapolate relationships to unsurveyed regions has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.

Model-based distance sampling has most commonly been implemented in a two-stage modelling framework.  In the first stage, detectability is estimated using the frequency of observed distances to fit the detection function.  In the second stage, detectability estimates from the first stage are used as an offset in a generalized additive model framework.  Detections are are binned into counts within user-defined sampling units (usually point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.  The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatial count model.   Due to the often sparse nature of wildlife survey data, this may require consideration of over- or under-dispersion and zero-inflated distributions and the negative-binomial or Tweedie distributions are common choices.  This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review.

A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}. Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  One-stage analyses of distance sampling data have tended to be Bayesian and use Markov chain Monte Carlo (MCMC) methods for inference, usually requiring data augmentation to model unobserved individuals or groups
% problem with this Royle reference here.  Not sure what I was intending to cite.
\citep{royle_HierarchicalModelSpatial_2008, schmidt_using_2012}.  \citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation and incorporate model uncertainty using reversible jump MCMC.

The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA \citep{rue_approximate_2009} and present an application on line transect distance sampling data.  \citet{yuan_point_2017} take a point process perspective and formulate the detection model as a thinning of a point pattern.  Key to their approach is formulating the detection model as the solution to an SPDE.  An approximate solution the SPDE is constructed using a B-spline basis and finite element methods which results in a sparse prior precision matrix for the detection model parameters which can be incorporated into the model in INLA.

A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, usually a key feature of detection functions.  The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. Representing the observation model as an SPDE also differs substantially from standard distance sampling models that use a parametric model for the detection function.  This may have hindered a wider adoption of using INLA for distance sampling analyses, \cite{yuan_point_2017} remains the only example we are aware of.  Furthermore, parametric forms of detection functions are readily extendible by including covariates and random effects to explain variation in detectability.  It is not clear how to do this the SPDE formulation.  These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA and the benefits of a one-stage modelling process.

Here we present an approach that is closely related to \citet{yuan_point_2017}.  We also fit the model using INLA and take advantage of the benefits the increased computational efficiency compared to MCMC.  However, we avoid the issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal detection function, as will be more familiar to users of classical distance sampling methods. However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \texttt{R-INLA} implementation of INLA \citep{rue_approximate_2009}.  We address this by using a method of iterated model fits based on a first-order Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.

We fit the model to point transect \akepa{} data, which differ from line transect methods in that there is a need to account for the increasing area surveyed as distance increases from the transect.  Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
